The function-sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-24 crelle-dictionary.tex
Use of Hom harmonized; improved wording in the introduction.

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\FF_q$; \'etale group schemes form a counterpoint to connected algebraic groups, since the component group of a smooth commutative group scheme is an \'etale group scheme. 
We find that the group homomorphism 
\begin{equation}
T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)
\end{equation}
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$, canonically, where $\Weil{} \subset \Gal({\bar\FF}_q/\FF_q)$ is the Weil group for $\FF_q$,
One can show, using the Kunneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
The fact that $T_G$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\FF_q)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.
Here we also see the importance of working with Weil sheaves; without them, we would not be able to geometrize the character of the \'etale group scheme $\ZZ$ 
with image $\ell^{\ZZ}$, for example. %given by $1\mapsto \ell$, for example.

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation} 
such as $A^*\ceq \Hom(A,\EEx)$ for any group $A$ and using $G$ for a smooth commutative group scheme over $\FF_q$,
in Section~\ref{sec:category} we define the category of quasicharacter sheaves on $G$.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to G(\FF_q)^*$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\FF_q)^* \rar & G(\Fq)^* \rar & G^0(\FF_q)^* \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

For any group $A$, we will write $A^*$ for the dual group $\Hom_\text{grp}(A, \EEx)$.

@@

\begin{proof}
  Recall the short exact sequence \eqref{eq:pi0}
  defining the component group scheme for $G$:
  \[
  \begin{tikzcd}
    1 \rar & G^0 \rar & G \rar & \pi_0(G) \rar & 0.
  \end{tikzcd}
  \]
  Let
  \[
  \begin{tikzcd}[row sep=20]
%  & 0 \arrow{d} & 0 \arrow{d} & 0\arrow{d} \\
    0 \arrow[dashed]{r} & \ker \TrFrob{\pi_0(G)} \arrow[dashed]{r} \dar & \arrow{d} \ker \TrFrob{G} \arrow[dashed]{r} & \ker \TrFrob{G^0} =0 \arrow[dotted, out=-10, in=170]{dddll} \dar & \\
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar \dar
    & \arrow{d} G(\Fq)^* \rar & G^0(\Fq)^* \rar \dar & 0\\
   &  \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \coker \TrFrob{G} \arrow[dashed]{r} & \coker \TrFrob{G^0} =0 & 
%    & \arrow{d} \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \arrow{d} \coker \TrFrob{G} \arrow[dashed]{r} &  \arrow{d} \coker \TrFrob{G^0} =0 & \\
%      & 0 & 0 & 0 \\
  \end{tikzcd}
  \]
  be the commutative diagram of abelian groups obtained by applying
  Lemma~\ref{lem:pullback} to \eqref{eq:pi0}.
 %
  The sequence of abelian groups
  \begin{equation}\label{eq:pi0k}
  \begin{tikzcd}
    1 \rar & G^0(\Fq) \rar & G(\Fq) \rar & \pi_0(G)(\Fq) \rar & 0,
  \end{tikzcd}
  \end{equation}
  is exact since $\Hh^1(\Fq,G^0) =0$ by Lemma~\ref{lem:G0alg-grp} and Lang's theorem on connected algebraic groups over finite fields \cite{lang:56a}.
  Since $\EEx$ is divisible, $\Hom_\text{Ab}(\ - \ ,\EEx)$ is exact and thus the dual sequence of
  character groups is also exact.
%
  The upper row is exact by Proposition~\ref{prop:middleexact}.
  Now Lemma~\ref{lem:G0alg-grp} and Proposition~\ref{prop:connected}
  imply that $\ker \TrFrob{G^0} =0$ and $\coker \TrFrob{G^0}=0$,
  while Proposition~\ref{prop:etale-iso} gives $\ker \TrFrob{\pi_0(G)} \iso \Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$
  and $\coker \TrFrob{\pi_0(G)}=0$.
%
It follows from the snake lemma that $\coker \TrFrob{G} =0$
and $\ker \TrFrob{\pi_0(G)} \to \ker \TrFrob{G}$ is an isomorphism.
This calculates $\ker  \TrFrob{G}$ and shows that
\[
  \begin{tikzcd}
0 \arrow{r} & \Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx)) \arrow{r} & \QCiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0 
  \end{tikzcd}
\]
is exact.
\end{proof}

@@

\begin{theorem}\label{thm:autornaught}
Let $G$ be a smooth commutative group scheme over $\Fq$.
%For all quasicharacter sheaves $\qcs{L}$ and $\qcs{L}'$ on $G$, every $\rho \in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism.
Moreover, 
\[
\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx),
\]
 canonically.
\end{theorem}
\begin{proof} 
%For the first part of the proposition, argue as in the analogous part of Proposition~\ref{prop:autornaught_etale}.
%For the second, f
Fix $\qcs{L} = (\gqcs{L},\mu,\phi)$ and consider the homomorphism $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ defined in Proposition~\ref{prop:autornaught_etale}.
This group homomorphism is clearly injective: if $\rho_x = \id_{\qcs{L}_x}$ for every $x\in \bG$ then $\rho = \id_{\qcs{L}}$.
Note that the image of $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ is contained in $\Hom_\text{cts}(\bG_{\Weil{}},\EEx)$, where $\bG_{\Weil{}}$ is given the topology of the base of the espace \'etal\'e attached to $\gqcs{L}$.
Since $\ell$ is invertible in $\Fq$, it follows that $\Hom_\text{cts}(\bG_{\Weil{}},\EEx) = \Hom(\pi_0(\bG_{\Weil{}}),\EEx)$. 
Note that $\pi_0(\bG_{\Weil{}})=\pi_0(\bG)_{\Weil{}}$. 
To see that $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ is surjective, begin with $\theta\in\Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and, for each $[x] \in \pi_0(\bG)_{\Weil{}}$ define $\bar\rho^y: \gqcs{L}^y \to \gqcs{L}^y$ by scalar multiplication by $\theta([x])\in \EEx$ for each $y\in [x]$.
This defines an isomorphism $\bar\rho : \gqcs{L}\to \gqcs{L}$ of local systems on $\bG$ compatible with $\mu$ and $\phi$, and thus an isomorphism $\rho :\qcs{L}\to \qcs{L}$ with maps to $\theta$ under $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$.
\end{proof}

>>>>

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\FF_q$; \'etale group schemes form a counterpoint to connected algebraic groups, since the component group of a smooth commutative group scheme is an \'etale group scheme. 
We find that the group homomorphism 
\begin{equation}
T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)
\end{equation}
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$, canonically, where $\Weil{} \subset \Gal({\bar\FF}_q/\FF_q)$ is the Weil group for $\FF_q$,
One can show, using the Kunneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
The fact that $T_G$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\FF_q)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.
Here we also see the importance of working with Weil sheaves; without them, we would not be able to geometrize any character of the \'etale group scheme $\ZZ$ 
with non-compact image, such as $\ell^{\ZZ}$.

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation} 
such as $A^*\ceq \Hom(A,\EEx)$ for any abelian group $A$ and using $G$ for a smooth commutative group scheme over $\FF_q$,
in Section~\ref{sec:category} we define the category of quasicharacter sheaves on $G$.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to G(\FF_q)^*$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20, column sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\FF_q)^* \rar & G(\Fq)^* \rar & G^0(\FF_q)^* \rar & 0,
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

@@

For any abelian group $A$, we will write $A^*$ for the dual group $\Hom(A, \EEx)$.

@@

\begin{proof}
  Recall the short exact sequence \eqref{eq:pi0}
  defining the component group scheme for $G$:
  \[
  \begin{tikzcd}
    1 \rar & G^0 \rar & G \rar & \pi_0(G) \rar & 0.
  \end{tikzcd}
  \]
  Let
  \[
  \begin{tikzcd}[row sep=20]
%  & 0 \arrow{d} & 0 \arrow{d} & 0\arrow{d} \\
    0 \arrow[dashed]{r} & \ker \TrFrob{\pi_0(G)} \arrow[dashed]{r} \dar & \arrow{d} \ker \TrFrob{G} \arrow[dashed]{r} & \ker \TrFrob{G^0} =0 \arrow[dotted, out=-10, in=170]{dddll} \dar & \\
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar \dar
    & \arrow{d} G(\Fq)^* \rar & G^0(\Fq)^* \rar \dar & 0\\
   &  \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \coker \TrFrob{G} \arrow[dashed]{r} & \coker \TrFrob{G^0} =0 & 
%    & \arrow{d} \coker \TrFrob{\pi_0(G)} =0 \arrow[dashed]{r} & \arrow{d} \coker \TrFrob{G} \arrow[dashed]{r} &  \arrow{d} \coker \TrFrob{G^0} =0 & \\
%      & 0 & 0 & 0 \\
  \end{tikzcd}
  \]
  be the commutative diagram of abelian groups obtained by applying
  Lemma~\ref{lem:pullback} to \eqref{eq:pi0}.
 %
  The sequence of abelian groups
  \begin{equation}\label{eq:pi0k}
  \begin{tikzcd}
    1 \rar & G^0(\Fq) \rar & G(\Fq) \rar & \pi_0(G)(\Fq) \rar & 0,
  \end{tikzcd}
  \end{equation}
  is exact since $\Hh^1(\Fq,G^0) =0$ by Lemma~\ref{lem:G0alg-grp} and Lang's theorem on connected algebraic groups over finite fields \cite{lang:56a}.
  Since $\EEx$ is divisible, $\Hom(\ - \ ,\EEx)$ is exact and thus the dual sequence of
  character groups is also exact.
%
  The upper row is exact by Proposition~\ref{prop:middleexact}.
  Now Lemma~\ref{lem:G0alg-grp} and Proposition~\ref{prop:connected}
  imply that $\ker \TrFrob{G^0} =0$ and $\coker \TrFrob{G^0}=0$,
  while Proposition~\ref{prop:etale-iso} gives $\ker \TrFrob{\pi_0(G)} \iso \Hh^0(\Weil{},\Hh^2(\pi_0({\bar G}),\EEx))$
  and $\coker \TrFrob{\pi_0(G)}=0$.
%
It follows from the snake lemma that $\coker \TrFrob{G} =0$
and $\ker \TrFrob{\pi_0(G)} \to \ker \TrFrob{G}$ is an isomorphism.
This calculates $\ker  \TrFrob{G}$ and shows that
\[
  \begin{tikzcd}
0 \arrow{r} & \Hh^0(\Weil{},\Hh^2(\pi_0(\bG),\EEx)) \arrow{r} & \QCiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0 
  \end{tikzcd}
\]
is exact.
\end{proof}

@@

\begin{theorem}\label{thm:autornaught}
Let $G$ be a smooth commutative group scheme over $\Fq$.
%For all quasicharacter sheaves $\qcs{L}$ and $\qcs{L}'$ on $G$, every $\rho \in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism.
Moreover, 
\[
\Aut(\qcs{L}) \iso \Hom(\pi_0(\bG)_{\Weil{}},\EEx),
\]
 canonically.
\end{theorem}
\begin{proof} 
%For the first part of the proposition, argue as in the analogous part of Proposition~\ref{prop:autornaught_etale}.
%For the second, f
Fix $\qcs{L} = (\gqcs{L},\mu,\phi)$ and consider the homomorphism $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ defined in Proposition~\ref{prop:autornaught_etale}.
This group homomorphism is injective: if $\rho_x = \id_{\qcs{L}_x}$ for every $x\in \bG$ then $\rho = \id_{\qcs{L}}$.
Note that homomorphisms in the image of $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ are continuous where $\bG_{\Weil{}}$ is given the topology of the base of the espace \'etal\'e attached to $\gqcs{L}$.
Since $\ell$ is invertible in $\Fq$, it now follows that the image of $\Aut(\qcs{L}) \to \Hom(\bG_{\Weil{}},\EEx)$ is contained in $\Hom(\pi_0(\bG_{\Weil{}}),\EEx)$. 
We also have $\pi_0(\bG_{\Weil{}})=\pi_0(\bG)_{\Weil{}}$. 
To see that $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ is surjective, begin with $\theta\in\Hom(\pi_0(\bG)_{\Weil{}},\EEx)$ and, for each $[x] \in \pi_0(\bG)_{\Weil{}}$ define $\bar\rho^y: \gqcs{L}^y \to \gqcs{L}^y$ by scalar multiplication by $\theta([x])\in \EEx$ for each $y\in [x]$.
This defines an isomorphism $\bar\rho : \gqcs{L}\to \gqcs{L}$ of local systems on $\bG$ compatible with $\mu$ and $\phi$, and thus an isomorphism $\rho :\qcs{L}\to \qcs{L}$ with maps to $\theta$ under $\Aut(\qcs{L}) \to \Hom(\pi_0(\bG)_{\Weil{}},\EEx)$.
\end{proof}
